---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- then run devtools::build_readme() to regenerate README.md.    -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# LakeEnsemblR.WQ

<!-- badges: start -->
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: GPL (>= 2)](https://img.shields.io/badge/License-GPL%20(%3E%3D%202)-blue.svg)](https://www.gnu.org/licenses/old-licenses/gpl-2.0.en.html)
<!-- badges: end -->

This package aims to facilitate the settin-up, running and calibrating and post-processing of the process based aquatic ecosystem models.
harmonize, and compare water quality simulations across multiple 1D lake model frameworks.

LakeEnsemblR.WQ is an extension of LakeEnsemblR package with a new framework that integrates water quality and biological modules for 1D process-based aquatic ecosystem models.
It aims to facilitate data preprocessing, setup, calibration, post-processing, and also support the calculation of system metrics through concept, state, process, system framework. 


LakeEnsemblR.WQ extends LakeEnsemblR workflows with tools to:

- Configure model-specific water quality and ecosystem dynamics settings,
- Validate model setups before run,
- Run model ensembles,
- Calibrate model ensembles -- Latin Hypercube sampling, optional Differential
  Evolution refinement
- Extract and harmonize output variables,
- Calculate standardized ecosystem metrics and summary statistics,
- Compare and visualize model outputs against each other and observations.

Supported 1D model frameworks:

- GLM-AED2
- GOTM-WET
- GOTM-Selmaprotbas
- Simstrat-AED2

## Installation

```r
# install.packages("remotes")
# Currently in development on tubabucak's fork; will move to the aemon-j
# organization (https://github.com/aemon-j/LakeEnsemblR.WQ) after release.
remotes::install_github("tubabucak/LakeEnsemblR.WQ")
```

Some dependencies are GitHub-only wrappers around external lake model
binaries: `GLM3r`, `WETr`, `SelmaprotbasR`, `SimstratR`, `MyLakeR`. These are
optional -- you only need the one(s) matching the model(s) you actually run.
If a required one is missing, `run_ensemble_wq()`/`run_lhc_wq()` will tell
you at call time rather than fail silently. All of them are listed in
`Remotes:` in `DESCRIPTION`, so `remotes::install_github()` installs them
automatically along with everything else.

`glmtools` is always required, no matter which model you run. It in turn
needs `GLM3r`, but doesn't say where to find it -- so tools that resolve
dependencies automatically (`pak`, and anything built on it, like
`devtools::install_deps()`/`build_vignettes()`) can fail with
`Can't find package called GLM3r`, even though `GLM3r` is listed in this
package's own `Remotes:`. If that happens, install `GLM3r` yourself first,
then install this package as normal:

```r
remotes::install_github("aemon-j/GLM3r@v3.3-lerwq")
remotes::install_github("tubabucak/LakeEnsemblR.WQ")
```

## Example projects

Full, real-lake example setups (config files, inputs, and run scripts) are
available at [tubabucak/LERWQ_testcases](https://github.com/tubabucak/LERWQ_testcases) --
useful as a starting template, or to see the full workflow applied end-to-end
against actual data.

## Typical workflow

```{r eval = FALSE}
library(LakeEnsemblR)

# 0) Export model-specific configuration and inputs for physical setup
# This step is only needed once per project, unless you want to change the
# physical configuration (e.g., bathymetry, inflows, meteorology, etc.).
# Please check the default configuration configuration that was generated before proceeding
# to the next step. You can edit the configuration file to change model settings if needed.
# For example the default configuration setup of GLM does not include the sediment module, so if you want to include the sediment module in GLM, you need to change the configuration file before proceeding to the next step.
# Or if you need more diagnostics and output variables, you can change the configuration file before proceeding to the next step.
export_config(
  config_file = "LakeEnsemblR.yaml",
  folder = ".",
  verbose = TRUE
)

library(LakeEnsemblR.WQ)


# 1) Export model-specific configuration and inputs for WQ setup
# This step is only needed once per project, unless you want to change the
# water quality configuration
export_config_wq(
  config_file = "LakeEnsemblR_WQ.yaml",
  folder = ".",
  verbose = TRUE
)

# 2) Run selected models
# This function only run the selected models with the current configuration setup
# and save their outputs in the output folders of the models.
run_res <- run_ensemble_wq(
  config_file = "LakeEnsemblR_WQ.yaml",
  models = c("GLM-AED2", "GOTM-WET", "GOTM-Selmaprotbas", "SIMSTRAT-AED2"),
  folder = ".",
  validate = TRUE,
  verbose = TRUE
)

# 3) cal_metric function computes the metrics (which was specified in Output.yaml) for the 
# selected models and save them in a list. Now they are all in the same units and can be compared with each other. 
# The metrics are computed for all the models that were run in the previous step.

metric_out <- cal_metrics(
  metric_yaml_file = "Output.yaml",
  model_filter = "all",
  wq_config_file = "LakeEnsemblR_WQ.yaml"
)

```

## Calibration

```{r eval = FALSE}

# Create calibration tables for all coupled models at once, with default bounds_factor = 0.2.
create_calibration_tables(
  folder = ".",
  config_file = "LakeEnsemblR_WQ.yaml",
  folder_out = "calibration",
  models_coupled = c("GLM-AED2", "GOTM-WET", "GOTM-Selmaprotbas", "Simstrat-AED2"),
  bounds_factor = 0.2
)

# Edit the generated calibration CSVs in the "calibration" folder before continuing:
#   - set include = TRUE for the parameters you want to calibrate
#   - adjust lower, upper, and initial as needed -- the generated CSVs also
#     include suggested min-max ranges (dict_min/dict_max) for some
#     parameters as a reference; use them to inform your own bounds if useful
#
# Then build a calibration setup table from the edited CSVs:
cs_all <- calib_setup_from_tables(
  folder_in = "calibration",
  model_coupled = c("GLM-AED2", "GOTM-WET", "GOTM-Selmaprotbas", "Simstrat-AED2")
)

# Calibrate all coupled models at once, with optional DE refinement and parallelism.
result_all <- cali_ensemble_wq(
  models          = c("GOTM-WET", "GOTM-Selmaprotbas", "GLM-AED2", "Simstrat-AED2"),
  calib_setup     = cs_all,
  yaml_file       = "Output.yaml",
  folder          = ".",
  n_samples       = 16,     # LHC seed
  use_de          = TRUE,
  de_popsize      = 12,
  de_iterations   = 8,
  de_parallel     = TRUE,   # <- must flip this on to use de_n_workers
  de_n_workers    = 4,
  n_workers       = 4,
  parallel_models = FALSE,   # keep this -- see why below
  obs_file        = "standart_observed_data.csv",
  wq_config_file  = "LakeEnsemblR_WQ.yaml",
  ler_config_file = "LakeEnsemblR.yaml",
  best_metric     = "KGE",
  target_variable   = "DO_gramsPerCubicMeter",
  parallel        = TRUE,
  force_parallel_glm_simstrat = TRUE,
  verbose         = TRUE,
  on_error        = "skip",
  save_results    = TRUE,
  output_dir      = "calibration_results",
  write_best      = TRUE,
  write_target    = "config"
)

`parallel_models` stacks with `parallel`/`de_parallel` (which parallelize LHC
samples or DE evaluations *within* one model's run) -- if you turn multiple of
these on together, size worker counts so
`n_model_workers * n_workers` (or `* de_n_workers`) doesn't oversubscribe your
CPU cores. Worker processes have no attached console, so with
`parallel_models = TRUE` progress messages are written to
`<output_dir>/cali_ensemble_wq_workers.log` instead of printing live -- tail
that file to watch progress while it runs.

result_all$summary                            # success/failure + row counts per model
result_all$best_parameter_sets[["GLM-AED2"]]   # winning parameters for one model
```

See the [Full Workflow](https://tubabucak.github.io/LakeEnsemblR.WQ/articles/full-workflow.html)
article (or `vignette("full-workflow")` if you installed with
`build_vignettes = TRUE`) for the full calibration workflow, including
single-model LHC/DE runs via `run_lhc_wq()`, writing best parameters back
with `write_best_calib_to_par_files()` and post-processing functions. See the
[Configuration File Reference](https://tubabucak.github.io/LakeEnsemblR.WQ/articles/config-reference.html)
article for a field-by-field breakdown of `LakeEnsemblR_WQ.yaml` and
`Output.yaml`.

## Key functions

- Setup and export:
  - export_config_wq()  is the main function to export the configuration files for the water quality models also calling the functions below to export the input files for the water quality models.
  - export_inputs()
  - set_up_configs()
  - set_value_config()
- Validation and run: These functions are used to validate the configuration files and input files for the water quality models before running the models. If the validation fails, the function will return an error message and stop the execution of the code. They are included in the export_config_wq() function and can also be called separately if needed.
  - validate_glm_aed()
  - validate_gotm_wet()
  - validate_simstrat()
  - run_ensemble_wq()
- Extraction and metrics: These functions are used to extract the output variables from the model runs and calculate the metrics for the water quality models. They are included in the cal_metrics() function and can also be called separately if needed.
  - get_output_wq()
  - extract_variable_list()
  - cal_metrics()
  - cal_stats()
  - export_all_stats()
- Calibration:
  - create_calibration_tables() / calib_setup_from_tables()
  - run_lhc_wq() -- single-model LHC sampling, with optional DE refinement
    (`use_de = TRUE`) and optional internal parallelism (`parallel`/`de_parallel`)
  - cali_ensemble_wq() -- calibrate multiple coupled models at once,
    sequentially or concurrently (`parallel_models = TRUE`)
  - write_best_calib_to_par_files() -- write the best parameter set back to
    calibration files
  - run_sensitivity() / run_multi_param_sensitivity()
- Compare and visualize:
  - compare_models_metric() / compare_models_metric_netcdf()
  - compare_plot() / scat_plot()
  - plot_heatmap_wq()
  - plot_strat_metrics() / plot_anoxic_metrics() / plot_ice_metrics()

## Input files

Common project files include:

- LakeEnsemblR_WQ master config (for model/module settings)
- LakeEnsemblR physical config
- Output metric config (for selected metric families)

Example from a typical setup:

- LakeEnsemblR_WQ.yaml
- LakeEnsemblR.yaml
- Output.yaml
- LakeEnsemblR_bathymetry_standard
- LakeEnsemblR_ice-height_standard
- LakeEnsemblR_inflow_standard
- LakeEnsemblR_meteo_standard
- LakeEnsemblR_outflow_standard

## Output structure

`cal_metrics()` returns a nested list by metric and model. For each metric/model combination, values are stored as data frames with `datetime` and depth/value columns.

Example access:

```{r eval = FALSE}
metric_out$GLM$TP_gramsPerCubicMeter
```

## Troubleshooting

If a model run succeeds but some metrics are missing:

1. Confirm the metric is enabled in Output.yaml.
2. Confirm the metric-variable mapping exists in the metrics dictionary.
3. Confirm the model actually wrote that variable to output.

For Simstrat-AED2 specifically, missing derived metrics (for example total chlorophyll-a or total phosphorus) are often caused by missing output variables in the Simstrat/AED2 output configuration (OutputDiagnosticVars should be 'true' ), not by run failure.

If you're developing against a local clone and see errors like
`unused arguments (...)` for a parameter that clearly exists in the
function's source, your session is likely running a stale **installed**
copy of the package rather than your edited source -- reload with
`devtools::load_all(".")` instead of `library(LakeEnsemblR.WQ)`.

## Development

```r
devtools::load_all(".")     # load current source (not a stale install)
devtools::document()        # regenerate NAMESPACE/man/ from roxygen tags
devtools::test()            # run the tests/testthat suite
devtools::check()           # full R CMD check
devtools::build_readme()    # regenerate README.md from this file
```

CI (`.github/workflows/R-CMD-check.yaml`) runs the same check on every push
and PR; `.github/workflows/pkgdown.yaml` rebuilds and publishes the
documentation site to the `gh-pages` branch.

## Contributing

1. Fork this repository.
2. Clone your fork.
3. Add the original repository as `upstream`.
4. Pull latest changes from `upstream/main`.
5. Create a feature branch, commit changes, and push.
6. Run `devtools::test()` (and ideally `devtools::check()`) before opening a PR.
7. Open a pull request.

Example git commands:

```bash
git clone git@github.com:<username>/LakeEnsemblR.WQ.git
cd LakeEnsemblR.WQ
git remote add upstream git@github.com:aemon-j/LakeEnsemblR.WQ.git
git fetch upstream
git checkout -b my-feature
git merge upstream/main
git push -u origin my-feature
```
