---
title: "LakeEnsemblR.WQ Full Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LakeEnsemblR.WQ Full Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Purpose

This vignette provides an end-to-end workflow for LakeEnsemblR.WQ:

1. Configure coupled model inputs.
2. Validate model setups.
3. Run model ensembles.
4. Compute harmonized metrics.
5. Prepare a calibration setup, run sensitivity analysis on it to identify
   which parameters actually matter before spending compute calibrating them,
   then calibrate -- single-model LHC, optional DE refinement, writing best
   parameters back, and calibrating all coupled models at once (sequentially
   or concurrently) with `cali_ensemble_wq()`.
6. Compare and visualize model outputs against each other and observations.

## Prerequisites

- A physical LakeEnsemblR setup already exported.
- A water-quality config file: LakeEnsemblR_WQ.yaml.
- A metrics config file: Output.yaml.
- Supported model folders, for example:
  - GLM-AED2
  - GOTM-WET
  - GOTM-Selmaprotbas
  - Simstrat-AED2
- For calibration against observations (Section 5): an observed-data CSV with
  columns `datetime`, `depth`, `variable_global_name`, `value`.

```{r eval = FALSE}
library(LakeEnsemblR)
library(LakeEnsemblR.WQ)
```

## 1) Export WQ configuration and inputs

```{r eval = FALSE}
# Export model-specific configuration and inputs for physical setup. This step is a must to create physical lake model folders
export_config("LakeEnsemblR.yaml")

# Export model-specific configuration and inputs for WQ setup
export_config_wq("LakeEnsemblR_WQ.yaml")


```

## 2) Validate model setups
These are to check if we have the correct model setup and configuration files for each model. If any of these fail, check the model folder and config files for errors.
```{r eval = FALSE}

validate_glm_aed(sim_folder = "GLM-AED2", file = "glm3.nml", verbose = TRUE)
validate_gotm_wet(sim_folder = "GOTM-WET", file = "gotm.yaml", verbose = TRUE)
validate_gotm_wet(sim_folder = "GOTM-SELMAPROTBAS", file = "gotm.yaml", verbose = TRUE)
validate_simstrat(sim_folder = "SIMSTRAT-AED2", file = "simstrat.par", verbose = TRUE)
```

`run_ensemble_wq()` (next section) can also run these automatically via its
own `validate = TRUE` argument -- calling them directly here is useful when
you want validation errors surfaced before committing to a full ensemble run.

## 3) Run ensemble simulations
### This function runs the ensemble of models with current configurations, and returns a list of run results. It also validates the model setups before running the models.
```{r eval = FALSE}
run_res <- run_ensemble_wq(
  config_file = "LakeEnsemblR_WQ.yaml",
  models = c("GLM-AED2", "GOTM-WET", "GOTM-Selmaprotbas", "SIMSTRAT-AED2"),
  folder = ".",
  validate = TRUE,
  verbose = TRUE
)
```

## 4) Compute metrics and statistics
This step can also be done after calibration to calculate the CSPS metrics or to see how the model behaves with default configuration. The metrics are calculated based on the Output.yaml file, which contains the list of metrics to be calculated for each model. The metrics are calculated for each model and stored in a list, which can be used for further analysis or visualization.

```{r eval = FALSE}
metric_out <- cal_metrics(
  metric_yaml_file = "Output.yaml",
  model_filter = "all",
  wq_config_file = "LakeEnsemblR_WQ.yaml"
)
```

## 5) Calibration workflow

### 5.1 Generate calibration tables

This function generates editable calibration CSVs for each model, based on
the parameter ranges defined in the configuration file. The CSVs can be
modified to adjust the parameter ranges for calibration. For some parameters,
literature ranges are also included in the CSVs to help you set up sensible
calibration ranges.

Pay attention if you have more than one phytoplankton or zooplankton group:
they share the same parameter names but have different `group_name` values.
Select which group to calibrate by setting `group_name` in the CSVs, and set
their ranges independently before running calibration. The calibration
tables are generated in the `"calibration"` folder, which can be edited and
then used for calibration. You may also want to adjust default values (N fixation, Silica module etc) in the model config files for diatom and cyanobacteria groups, as they are used as initial values for calibration.

```{r eval = FALSE}
create_calibration_tables(
  folder = ".",
  config_file = "LakeEnsemblR_WQ.yaml",
  folder_out = "calibration",
  models_coupled = c("GLM-AED2", "GOTM-WET", "GOTM-Selmaprotbas", "SIMSTRAT-AED2"),
  bounds_factor = 0.2
)

```

Edit module files in the calibration folder:
- set include = TRUE for selected parameters
- adjust lower, upper, and initial as needed

### 5.2 Build calib_setup from edited tables

```{r eval = FALSE}
cs_all <- calib_setup_from_tables(
  folder_in = "calibration",
  model_coupled = c("GLM-AED2", "GOTM-WET", "GOTM-Selmaprotbas", "SIMSTRAT-AED2")
)
```

### 5.3 Sensitivity analysis

Before committing to a full LHC/DE calibration run, it's worth checking
which parameters in `cs_all` actually move the outputs you care about --
sensitivity analysis on the calibration setup you just built, before
spending compute on calibrating parameters that turn out not to matter (or
whose ranges are far wider than they need to be).

Two functions cover this, both supporting all four coupled models
(`"GLM-AED2"`, `"GOTM-WET"`, `"GOTM-Selmaprotbas"`, `"Simstrat-AED2"`) via the
same `model` argument used elsewhere in this vignette:

- `run_sensitivity()` -- **one parameter at a time**: sweeps a single
  parameter across `n_steps` values between its `calib_setup` bounds, holding
  everything else fixed at whatever is currently in the model's config files.
- `run_multi_param_sensitivity()` -- **all parameters together**: at each of
  `n_steps` iterations, every parameter in `param_names` (default: every row
  of `calib_setup`) is sampled jointly via Latin Hypercube Sampling and varied
  simultaneously. 

Both support two output modes:

- `output_mode = "raw"` -- bypasses the metrics dictionary/`Output.yaml`
  entirely and pulls one or more variables straight from the model's native
  output via `get_output_wq()`. Requires `vars` (the model-native variable
  name, e.g. `"selmaprotbas_DO_mg"` for GOTM-Selmaprotbas, `"OXY_oxy"` for
  GLM-AED2/Simstrat-AED2, `"sO2W"` for GOTM-WET) instead of `wq_config_file`.
  Useful for a quick look at a raw variable without needing a dictionary
  entry for it.
- `output_mode = "metrics"` (default) -- runs `cal_metrics()` at each step,
  same as calibration/Section 4. Requires `wq_config_file` and a
  `target_variable` (optional) naming which metric(s) in `Output.yaml` you
  care about, e.g. `"DO_gramsPerCubicMeter"`. But it takes longer to run because it computes every metric in `Output.yaml` for each


#### 5.3.1 One-at-a-time: `run_sensitivity()`

```{r eval = FALSE}
cs_selma <- subset(cs_all, model_coupled == "GOTM-Selmaprotbas")

# raw mode -- no wq_config_file needed, just the model-native variable name
res_sens <- run_sensitivity(
  param_name  = "kc",                       # light extinction coefficient
  calib_setup = cs_selma,
  yaml_file   = "Output.yaml",
  model_dir   = "GOTM-Selmaprotbas",
  n_steps     = 10,
  model       = "GOTM-Selmaprotbas",
  output_mode = "raw",
  vars        = "selmaprotbas_DO_mg"
)

```

`results[[i]]$param_value` gives the parameter value; `results[[i]]$output`
(raw mode) or `results[[i]]$metrics` (metrics mode) gives that step's result
-- see `?run_sensitivity` for the exact nested shape of each.

##### Phytoplankton/zooplankton group parameters

Parameters shared by multiple groups (e.g. `r0`, present in both `diatoms`
and `cyanobacteria`) produce multiple `calib_setup` rows with the same `pars`
but different `group_name`. Pass `group_name` to target just one group;
omitting it sweeps **all** matching groups' rows with the same value,
simultaneously, every iteration -- which is rarely what you want unless
that's deliberate:

```{r eval = FALSE}
res_sens_diatoms <- run_sensitivity(
  param_name  = "r0",
  calib_setup = cs_selma,
  yaml_file   = "Output.yaml",
  model_dir   = "GOTM-Selmaprotbas",
  n_steps     = 10,
  model       = "GOTM-Selmaprotbas",
  group_name  = "diatoms",                  # only this group's r0 is varied
  output_mode = "raw",
  vars        = "selmaprotbas_DO_mg"
)
```

#### 5.3.2 Joint multi-parameter sensitivity: `run_multi_param_sensitivity()`

```{r eval = FALSE}
res_multi <- run_multi_param_sensitivity(
  # param_names left NULL -> uses every row of cs_selma$pars in the table
  calib_setup = cs_selma,
  yaml_file   = "Output.yaml",
  model_dir   = "GOTM-Selmaprotbas",
  n_steps     = 20,
  model       = "GOTM-Selmaprotbas",
  output_mode = "raw",
  vars        = "selmaprotbas_DO_mg"
)
```

Bounds come from `calib_setup$lb`/`ub` by default; pass `rel_change` (e.g.
`0.1` for +/-10%) instead to sample symmetrically around `calib_setup$x0`.
Duplicate `pars` names (group parameters) in `param_names` are matched
positionally to `calib_setup`'s own row order for that name -- check
`cs_selma[cs_selma$pars == "r0", c("pars", "group_name")]` first if the order
isn't obvious.

#### 5.3.3 Visualizing an uncertainty envelope

For `output_mode = "raw"` with a depth-resolved variable, stack every
iteration into long format and summarize a band (e.g. a 90% interval) plus a
central line per timestep:

```{r eval = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

stack_sensitivity_raw <- function(res, depth = NULL) {
  pieces <- lapply(seq_along(res), function(i) {
    df <- res[[i]]$output
    if (is.list(df) && !is.data.frame(df)) df <- df[[1]]
    if (is.null(df) || !is.data.frame(df) || nrow(df) == 0) {
      warning("Iteration ", i, " (param_value = ", res[[i]]$param_value,
              ") has no usable output -- skipping.")
      return(NULL)
    }
    long <- df %>%
      pivot_longer(cols = starts_with("Depth_"), names_to = "depth_col", values_to = "value") %>%
      mutate(depth = as.numeric(sub("Depth_", "", depth_col)))
    if (!is.null(depth)) {
      nearest <- long$depth[which.min(abs(long$depth - depth))]
      long <- long[long$depth == nearest, ]
    }
    long$iteration <- i
    long
  })
  do.call(rbind, pieces)
}

df_long <- stack_sensitivity_raw(res_sens_diatoms, depth = 5)

df_env <- df_long %>%
  group_by(datetime) %>%
  summarise(
    q_lo   = quantile(value, 0.05, na.rm = TRUE),
    q_hi   = quantile(value, 0.95, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  geom_line(data = df_long, aes(datetime, value, group = iteration),
            color = "grey70", alpha = 0.25, linewidth = 0.3) +
  geom_ribbon(data = df_env, aes(datetime, ymin = q_lo, ymax = q_hi),
              fill = "steelblue", alpha = 0.25) +
  geom_line(data = df_env, aes(datetime, median), color = "steelblue", linewidth = 0.8) +
  labs(x = NULL, y = "DO (mg/m3)", title = "Sensitivity envelope: kc -> DO at 5 m") +
  theme_minimal()
```



### 5.4 Run small LHC test per model

Run one model at a time. `run_lhc_wq()` samples the parameter space with Latin
Hypercube Sampling and, for each sample, edits the model's own config files in
`model_dir` in place, runs the model, and scores it. If `obs_file` is supplied
it returns a long-format data frame (one row per iteration x observed
variable) with NSE/RMSE/NRMSE/PBIAS/KGE; otherwise it is
`cal_metrics()`-derived metrics only.

```{r eval = FALSE}
cs_glm <- subset(cs_all, model_coupled == "GLM-AED2")

param_glm <- unique(cs_glm$pars)
res_glm <- run_lhc_wq(
  model          = "GLM-AED2",
  param_names    = param_glm,
  calib_setup    = cs_glm,
  yaml_file      = "Output.yaml",
  model_dir      = "GLM-AED2",
  n_samples      = 10,
  wq_config_file = "LakeEnsemblR_WQ.yaml",
  obs_file       = "standart_observed_DO.csv",   # datetime, depth, variable_global_name, value
  best_metric    = "KGE",
  verbose        = TRUE
)

# Note: model_dir's config files are restored to their pre-calibration state
# once run_lhc_wq() returns (including on error) -- the function snapshots
# model_dir before sampling and copies that snapshot back via on.exit(). Your
# working files are not left mutated at the last-tried parameter set.

head(res_glm)
attr(res_glm, "best_parameter_set")   # winning parameter set for this model
```



### 5.5 Optional: refine with Differential Evolution (DE)

Set `use_de = TRUE` to run DEoptim after the LHC phase, seeded from the best
LHC results. When DE is used, `attr(res_glm, "best_parameter_set")` is
overwritten with DE's refined optimum (DE improves on the LHC seed) -- the
original LHC-phase best is still available under
`attr(res_glm, "lhc_best_parameter_set")` for comparison.

```{r eval = FALSE}
    res <- run_lhc_wq(
        model          = "GLM-AED2",
        param_names    = param_glm,
        calib_setup    = cs_glm,
        yaml_file      = "Output.yaml",
        model_dir      = "GLM-AED2",
        obs_file       = "standart_observed_DO.csv",
        wq_config_file = "LakeEnsemblR_WQ.yaml",
        verbose        = TRUE,
        save_results   = FALSE,
        stats_by_depth = TRUE,
        return_best    = TRUE,
        best_metric    = "KGE",
        
        parallel       = TRUE,
        n_workers      = 2,
        n_samples      = 10,
        
       use_de         = TRUE,
       de_parallel    = TRUE,
       de_n_workers   = 2,
       de_popsize     = 4,
       de_iterations  = 1,
      target_variables = c("DO_gramsPerCubicMeter")
    )

attr(res, "best_parameter_set")   # DE-refined best (used for write-back below)

```

### 5.6 Write the best parameters back

`write_best_calib_to_par_files()` takes the best row (LHC's, or DE's if
`use_de = TRUE`) and writes it into the module `par_file` CSVs referenced by
`LakeEnsemblR_WQ.yaml` (or directly into model config files with
`write_target = "config"`). Re-run `export_config_wq()` afterward if you need
those values pushed all the way into the model folders for a final
verification run.

```{r eval = FALSE}
write_best_calib_to_par_files(
  lhc_results  = res_glm,
  calib_setup  = cs_glm,
  config_file  = "LakeEnsemblR_WQ.yaml",
  folder       = ".",
  metric       = "KGE",
  write_target = "config"
)

```



Re-run the model so output.nc reflects those parameters


```{r eval = FALSE}
run_ensemble_wq(
  config_file = "LakeEnsemblR_WQ.yaml",
  models = c("GLM-AED2"),
  folder = ".",
  validate = TRUE,
  verbose = TRUE
)

```


### 5.7 Calibrate all coupled models with cali_ensemble_wq()

`cali_ensemble_wq()` is the multi-model wrapper around `run_lhc_wq()`: pass it
the full `models` vector and the combined `cs_all` table (split automatically
by `model_coupled`) and it runs, scores, and optionally writes back the best
parameters for every model in one call.

By default the four models run sequentially. Since each model has its own
`model_dir`, they can safely run **concurrently** instead, one worker process
per model:
Start with small number of iterations to see if the models run successfully, then increase the number of iterations and workers for a full calibration run.

```{r eval = FALSE}

result_all <- cali_ensemble_wq(
  models          = c("GOTM-WET", "GOTM-Selmaprotbas"),
  calib_setup     = cs_all,
  yaml_file       = "Output.yaml",
  folder          = ".",
  n_samples       = 4,
  obs_file        = "standart_observed_DO.csv",
  wq_config_file  = "LakeEnsemblR_WQ.yaml",
  ler_config_file = "LakeEnsemblR.yaml",
  best_metric     = "KGE",
  parallel        = TRUE,
  force_parallel_glm_simstrat = TRUE,
  parallel_models = FALSE,
  use_de          = TRUE,
  de_parallel     = TRUE,
  de_n_workers    = 4,
  de_popsize      = 4,
  de_iterations   = 3,
  verbose         = TRUE,
  on_error        = "skip",
  save_results    = TRUE,
  output_dir      = "calibration_results",
  write_best      = TRUE,
  write_target    = "config"
)
```

`parallel_models` stacks with `parallel`/`de_parallel` (which parallelize LHC
samples or DE evaluations *within* one model's run) -- if you turn multiple of
these on together, size worker counts so
`n_model_workers * n_workers` (or `* de_n_workers`) doesn't oversubscribe your
CPU cores. Worker processes have no attached console, so with
`parallel_models = TRUE` progress messages are written to
`<output_dir>/cali_ensemble_wq_workers.log` instead of printing live -- check
that file to watch progress while it runs.

## 6) Compare and visualize model outputs

The functions below all plot directly into outputs you've already produced
earlier in this vignette: `cal_metrics()`'s nested list (Section 4) and the
NetCDF file from `create_netcdf_output()`.

### 6.1 Compare one metric across models

`compare_models_metric()` takes the list returned by `cal_metrics()` directly
and line-plots one metric across all coupled models (auto-detecting the
depth/value columns from that structure).

```{r eval = FALSE}
metric_out <- cal_metrics(
  metric_yaml_file = "Output.yaml",
  model_filter = "all",
  wq_config_file = "LakeEnsemblR_WQ.yaml"
)

cmp <- compare_models_metric(
  metric_out = metric_out,
  metric     = "Temp_degreeCelcius",
  depth      = 1
)
cmp$plot
```

`compare_models_metric_netcdf()` does the same comparison but reads straight
from a NetCDF file (see 7.3) instead of a `cal_metrics()` list 

```{r eval = FALSE}
cmp_nc <- compare_models_metric_netcdf(
  nc_file = "output.nc",
  metric  = "Temp_degreeCelcius",
  depth   = 1
)
cmp_nc$plot
```

### 6.2 Stratification metrics across models and years

`plot_strat_metrics()` accepts the same `cal_metrics()` list (or a NetCDF
path) and produces three plots per model: stratification duration, onset day
of year, and mixing onset day of year.

```{r eval = FALSE}
strat <- plot_strat_metrics(metric_out, metric_name = "Duration_of_Stratification")
strat$duration_plot
strat$onset_plot
strat$mixing_plot
```

### 6.3 Depth-time heatmap from the ensemble NetCDF

`create_netcdf_output()` combines all coupled models' output into a single
NetCDF for easier downstream analysis; `plot_heatmap_wq()` reads it directly
to produce a depth-time heatmap faceted by model. Repeat the `plot_heatmap_wq()`
call for any metric name present in `Output.yaml` (e.g. `DO_gramsPerCubicMeter`,
`NH4_gramsPerCubicMeter`, `TP_gramsPerCubicMeter`, `Total_Chla_miligramsPerCubicMeter`).

```{r eval = FALSE}
create_netcdf_output(
  output_lists    = metric_out,
  folder          = ".",
  model           = c("GLM", "SELMAPROTBAS", "WET", "SIMSTRAT"),
  ler_config_file = "LakeEnsemblR.yaml",
  wq_config_file  = "LakeEnsemblR_WQ.yaml",
  out_file        = "ensemble_output.nc"
)

plot_heatmap_wq("ensemble_output.nc", "Temp_degreeCelcius", spin_up = 30)
```

### 6.4 Anoxic and ice metrics across models and years

`plot_anoxic_metrics()` and `plot_ice_metrics()` follow the same flexible
input as `plot_strat_metrics()` (7.2) -- a `cal_metrics()` list or a NetCDF
path -- and plot their respective derived metrics per model per year.

```{r eval = FALSE}
plot_anoxic_metrics("ensemble_output.nc")
plot_ice_metrics("ensemble_output.nc", metric_name = "Ice_Thickness_meter")
```


### 6.5 Plot the best parameters (main config folder)

You can check the dictionary for the global variable name of the observed data, and use it to plot the model vs observed data. The variable_global_name should match the variable_global_name column in the observed data file.
```{r eval = FALSE}
result <- plot_model_vs_obs_wq(
    config_file           = "Output.yaml",
    model                 = "GLM-AED2",
    vars                  = "temp",              # GLM-native variable name
    obs_data              = "standart_observed_data.csv",
    variable_global_name  = "Temp_degreeCelcius", # matches obs_data's variable_global_name column
    y_title               = "DO (mmol/m3)"
)

result <- plot_model_vs_obs_wq(
    config_file           = "Output.yaml",
    model                 = "GLM-AED2",
    vars                  = "OXY_oxy",              # GLM-native variable name
    obs_data              = "standart_observed_data.csv",
    variable_global_name  = "DO_gramsPerCubicMeter", # matches obs_data's variable_global_name column
    y_title               = "DO (g/m3)"
)

# vars can be omitted -- it's then auto-derived from the metrics dictionary
# using model + variable_global_name, which also works for the other coupled
# models (not just GLM-AED2).
result <- plot_model_vs_obs_wq(
    config_file           = "Output.yaml",
    model                 = "GOTM-Selmaprotbas",
    obs_data              = "standart_observed_data.csv",
    variable_global_name  = "Total_Chla_miligramsPerCubicMeter", # matches obs_data's variable_global_name column
    y_title               = "Chl-a (mg/m3)"
)

```

### 6.6 Per-model time series and scatter plots against observations

`compare_plot()` and `scat_plot()` work at a lower level than the functions
above: instead of a `cal_metrics()` list, each takes one **wide-format**
data frame per model (`datetime` + `Depth_<n>` columns -- the shape
`get_output_wq()` returns) plus one for observations, and compute
`cal_stats()` against the observed values internally. They're most useful
once you already have per-model output pulled out via `get_output_wq()`
(see `?compare_plot` and `?scat_plot` for the full column/argument
requirements) -- for example:

```{r eval = FALSE}
data_glm <- get_output_wq(
  config_file = "Output.yaml",
  model       = "GLM",
  vars        = "temp",
  depth_01    = 1
)[[1]]

# ...similarly for data_wet, data_selma, data_simstrat, and data_obs

res <- compare_plot(
  data_glm, data_wet, data_selma, data_simstrat, data_obs,
  depth   = 1,
  y_title = "Temperature (°C)"
)
res[[1]]        # the plot
res[[2]]        # cal_stats() output for GLM at this depth
```

